Describing the scale and composition of calls for police service: a replication and extension using open data

### 2022-02-10 paper.Rmd
Edited until Detroit data

---
title: 'Describing the scale and composition of calls for police service: a replication and extension using open data'
author: Samuel Langton\dag, Stijn Ruiter\dag, Tim Verlaan\dag
bibliography: refs.bib
output:
  pdf_document: 
    number_sections: true
header-includes:
  - \usepackage{lscape}
  - \usepackage{lineno}
  - \linenumbers
csl: apa.csl
abstract: "This paper describes the scale and composition of emergency demand for police services in Detroit, United States. The contribution is made in replication and extension of analyses reported by Ratcliffe (2021) and Lum et al. (2021). Findings indicate that police spend a considerable proportion of time performing a social service function. Just 47% of total deployed time responding to 911 calls is consumed by crime incidents. The remainder is spent on community issues (6%), health (7%), proactive (5%), quality of life (19%) and traffic duties (13%). A small number of calls consume a disproportionately large amount of police officer time. Emergency demand is concentrated in time and space, and can differ between types of demand. Findings further highlight the potential implications of radically reforming police forces in the United States. The data and code used here are openly available for reproduction, re-use and scrutiny."
---

@@

Keywords: police, calls for service, 911, demand, reactive.

@@

\dag Netherlands Institute for the Study of Crime and Law Enforcement, Amsterdam, Netherlands. \linebreak \linebreak \linebreak \linebreak

@@

**Disclosure statement** The author declares that they have no competing interests.

**Funding details** To be completed.

@@

A recent study from @ratcliffe2021policing sought to describe the complexity and diversity of public demand for police services in Philadelphia, United States (US), during 2019. The study was motivated by the recognition that police spend a considerable proportion of deployed resource resolving incidents which fall outside of the traditional crime-fighting role of the police. Instead, the origins of public demand for the police can often be attributed to a lack of supply and/or accessibility failure in other organisations, such as those providing support for people requiring (mental) health assistance [@wood2021can; @van2017law]. Ratcliffe reported that just 40% of public emergency calls for service during the year were crime incidents. These calls consumed 55% of total officer shift time. The remaining public demand from calls involved community issues (10% of calls, 7% of time), medical/public health incidents (7% of calls, 9% of time), proactive policing (8% of calls, 5% of time), quality of life (24% of calls, 14% of time) and traffic duties (11% of calls and 11% of time). Comparable results were subsequently reported in the US across multiple anonymized police jurisdictions [@lum2021can; see also @lumwash in _The Washington Post_].

@@

Here, we propose an extension of @ratcliffe2021policing and @lum2021can. This extension has two principal components. First, we replicate the main analyses of these existing investigations into the scale and composition of police demand using open study materials. To date, investigations have been conducted on data which is not publicly available and without the materials (e.g., code) to reproduce the results. Open science practices have been identified as fundamental to the aims of evidence-based policing and vital in ensuring that in the long-term, real-life decisions are based on a concrete and replicable pool of evidence [@openedp2019]. In this case, the necessity for a fully open investigation into the scale and composition of police demand is particularly important given the public nature of debate around police reform and the considerable implications of redefining the role of the police without transparent empirical investigation. It is reasonable for police practitioners and the public to expect a thorough, open and replicable investigation into police demand before academia concludes that the issue is resolved – certainly before findings are used to change or introduce police practice or public policy. Open materials have the additional benefit of being available for re-use in other jurisdictions, helping build a reliable and comparable evidence-base around the scale and composition of police demand across different contexts.

@@

Second, we extend the analyses used by @ratcliffe2021policing and @lum2021can. This includes a direct comparison between different measures of 'officer time', namely, total deployment time and time on scene. These measures have been reported separately in each existing analyses, so the precise impact of their usage on the same data is unclear. We also provide a comprehensive account of the variability in the time consumed by different demand classifications (e.g., crime, community, health). These statistics shed light on the degree to which a disproportionately large amount of officer time is consumed by just a small number of calls - insight which can be used to assess how a reduction in demand, or a shift away from calls for police service towards other social services, might free-up officer time. In addition, we report on the spatial and temporal patterning of each demand classification. Distinctions in these patterns across demand types would be indicative of similarly distinct demand-generating mechanisms (e.g,, environmental characteristics). Such concentrations in time and space might also inform deployment decisions based on officer expertise and experience, and/or collaboration with other services (e.g., mental health professionals).

@@

The spatio-temporal patterning of this demand was summarized in two visuals, focusing specifically on those calls relating to medical/public health. A heatmap showed the mean call volume during hours of the day and days of the week, identifying peak times for demand. A geographic map visualized the spatial patterning of these call count concentrations using kernel density in order to identify hotspots of police demand involving medical/public health issues. To supplement these findings, Ratcliffe also reported on the differences between the initial CAD incident type classification and the final disposition using a sankey diagram -- again focusing specifically on medical and public health calls.

>>>>

---
title: 'Describing the scale and composition of calls for police service: a replication and extension using open data'
author: Samuel Langton\dag, Stijn Ruiter\dag\*, Tim Verlaan\dag
bibliography: refs.bib
output:
  pdf_document: 
    number_sections: true
header-includes:
  - \usepackage{lscape}
  - \usepackage{lineno}
  - \linenumbers
csl: apa.csl
abstract: "This paper describes the scale and composition of emergency demand for police services in Detroit, United States. The contribution is made in replication and extension of analyses reported by Ratcliffe (2021) and Lum et al. (2021). Findings indicate that police spend a considerable proportion of time performing a social service function. Just 47% of total deployed time responding to 911 calls is consumed by crime incidents. The remainder is spent on community (6%), health (7%), proactive (5%), quality of life (19%), and traffic (13%) duties. A small number of calls consume a disproportionately large amount of police officer time. Emergency demand is concentrated in time and space, and can differ between types of demand. Findings further highlight the potential implications of radically reforming police forces in the United States. The data and code used here are openly available for reproduction, re-use and scrutiny."
---

@@

Keywords: police, calls for service, 911, demand

@@

\dag Netherlands Institute for the Study of Crime and Law Enforcement (NSCR), Amsterdam, Netherlands. \linebreak
\* Department of Sociology, Utrecht University, Utrecht, Netherlands \linebreak \linebreak \linebreak \linebreak

@@

**Disclosure statement** The authors declare that they have no competing interests.

**Funding details** This research is part of the NSCR research program _What Works in Policing. Towards	Evidence-Based Policing in the Netherlands_, which has been funded by Netherlands Police.

@@

A recent study from @ratcliffe2021policing sought to describe the complexity and diversity of public demand for police services in Philadelphia, United States (US), during 2019. The study was motivated by the recognition that police spend a considerable proportion of deployed resource resolving incidents which fall outside of the traditional crime-fighting role of the police. Instead, the origins of public demand for the police can often be attributed to a lack of supply and/or accessibility failure in other organisations, such as those providing support for people requiring (mental) health assistance [@wood2021can; @van2017law]. Ratcliffe reported that just 40% of public emergency calls for service during the year were crime incidents. These calls consumed 55% of total officer shift time. The remaining public demand from calls involved community issues (10% of calls, 7% of time), medical/public health incidents (7% of calls, 9% of time), proactive policing (8% of calls, 5% of time), quality of life (24% of calls, 14% of time) and traffic duties (11% of calls and 11% of time). Comparable results were subsequently reported across multiple anonymized police jurisdictions in the US [@lum2021can; see also @lumwash in _The Washington Post_].

@@

Here, we propose an extension of @ratcliffe2021policing and @lum2021can. This extension has two principal components. First, we replicate the main analyses of these existing investigations into the scale and composition of police demand using open study materials. To date, investigations have been conducted on data which are not publicly available and without the materials (e.g., code) to reproduce the results. Open science practices have been identified as fundamental to the aims of evidence-based policing and vital in ensuring that in the long-term, real-life decisions are based on a concrete and replicable pool of evidence [@openedp2019]. In this case, the necessity for a fully open investigation into the scale and composition of police demand is particularly important given the public nature of the debate around police reform and the considerable implications of redefining the role of the police without transparent empirical investigation. It is reasonable for police practitioners and the public to expect a thorough, open and replicable investigation into police demand before academia concludes that the issue is resolved – certainly before findings are used to change or introduce police practice or public policy. Open materials have the additional benefit of being available for re-use in other jurisdictions, helping build a reliable and comparable evidence-base around the scale and composition of police demand across different contexts.

@@

Second, we extend the analyses used by @ratcliffe2021policing and @lum2021can. This includes a direct comparison between different measures of 'officer time', namely, total deployment time and time on scene. These measures have been reported separately in the existing analyses, so the precise impact of their usage on the same data is unclear. We also provide a comprehensive account of the variability in the time consumed by different demand classifications (e.g., crime, community, health). These statistics shed light on the degree to which a disproportionately large amount of officer time is consumed by just a small number of calls - insight which can be used to assess how a reduction in demand, or a shift away from calls for police service towards other social services, might free-up officer time. In addition, we report on the spatial and temporal patterning of each demand classification. Distinctions in these patterns across demand types would be indicative of similarly distinct demand-generating mechanisms (e.g., built environment and sociodemographic backdrop). Such concentrations in time and space might also inform deployment decisions based on officer expertise and experience, and/or collaboration with other services (e.g., mental health professionals).

@@

The spatio-temporal patterning of this demand was summarized in two visuals, focusing specifically on those calls relating to medical/public health. A heatmap showed the mean call volume during hours of the day and days of the week, identifying peak times for demand. A geographic map visualized the spatial patterning of these call count concentrations using kernel density estimation in order to identify hotspots of police demand involving medical/public health issues. To supplement these findings, Ratcliffe also reported on the differences between the initial CAD incident type classification and the final disposition using a Sankey diagram -- again focusing specifically on medical and public health calls.
